Deep Reinforcement Learning for Conservation Decisions

### 2021-05-13 manuscript/manuscript.Rmd
edits to ethics/power paragraph

Any application of RL can be divided into two components: an *environment* and an *agent*.
An *environment* is typically a computer simulation, though it is possible to use the real world as the RL environment [@robotics].
Over the course of a task, an *agent*, which is often a computer program, continuously interacts with the environment. 
At each time step, the agent observes the current *state* of the environment then performs an *action*; as a result of this action, the environment will transition to a new state and will transmit a numerical *reward* signal to the agent.
The goal of the agent is to learn how to maximize its expected cumulative reward.
The agent learns how to achieve this objective during a period called *training*.
In training, the agent *explores* the available states and actions. 
Once the agent comes across a highly rewarding sequence of states and actions, the agent will reinforce this behavior so that it is more likely for the agent to *exploit* the same high reward trajectory in the future.
Throughout this process, the agent's behavior is codified into what is called a *policy*, which describes what action an agent should take for a given state.

To solve the RL problem, algorithms either take a *model-free* or *model-based* approach. 
The distinction is that *model-free* algorithms do not attempt to learn or use a model of the environment.
Conversely, *Model-based* algorithms either learn or use a model of the environment to achieve the RL objective.
A trade-off between these approaches is that when it is possible to quickly learn a model of the environment or the model is already known, model-based algorithms tend to require much less interaction with the environment to learn good-performing policies [@mbpo].
Yet, frequently, learning a model of the environment is very difficult, and in these cases, model-free algorithms tend to outperform [@CITE].
Since most of the recent focus in RL has gone towards model-free algorithms, we will focus on model-free agents for the rest of the paper.

Among model-free algorithms, there is divergence around what functions the agent is attempting to learn in order to achieve the RL objective.
Generally, model-free agents are either learning a policy function, a value function or both.
For context, *value functions* are proxies for how high of a cumulative reward an agent can expect to receive from a given state or state-action pair.
If an agent knows the value function, then the agent can find the optimal action at any state by selecting the action that will maximize the value function in expectation.
The class of algorithms that exclusively try to learn a value function are called *value learning* methods.
*Policy gradient* algorithms exclusively learn the policy.
And, lastly, *actor-critic* algorithms attempt to learn both a policy function and a value function, whereby the value function, which is learned separately from the policy, is used to inform the agent on the goodness of a selected action.
There is a handful of advantages and disadvantages for each of these approaches that we will not go in depth on -- for more, see [@CITE]; yet, the main point of divergence comes from the bias-variance trade-off [@CITE].
Generally, policy gradient algorithms tend to struggle with high variance, value-based learning tends to be biased while actor-critic methods balance these two concerns [@CITE; @CITE; @CITE].

Neural networks become useful in RL when the environment has a large state-action space, which happens frequently with realistic decision-making problems. 
Classic RL algorithms, like policy iteration, value iteration and Actor-Critic [@suttonbarto], fail when the state-action space is large, e.g. when the possible states and actions can not be represented in a tractable table [@CITE].
Yet, over the last 10 years, researchers have shown that neural network-based RL algorithms can perform well on previously unsolvable environments.
Neural networks have been widely useful in reinforcement learning because neural nets have the property of being general function approximators [@CITE].
Whenever there is a need for an agent to approximate some function, say a policy function, value function or a transition function, neural networks can be used in this capacity.
While there are other function approximators that can be used in RL, e.g. Gaussian processes, neural networks have excelled in this role because of their ability to learn complex, non-linear, high dimensional functions and their ability to adapt given new information [@CITE].
We will not discuss in detail how neural networks work -- see [@CITE] for further background --, but the general process for how neural networks are used in RL is that at each learning step, the agent, say a policy gradient-based agent, will adjust the parameters in its policy neural network so that according to past experiences, the agent will be more likely to engage in highly rewarding behavior in the future. 
This same process occurs in value learning and actor-critic-based deep RL algorithms, except these agents are adjusting either a value network or both a value and policy network respectively.

@@

An environment is a mathematical function, computer program, or real world experience that takes an agent's proposed *action* as input and returns an observation of its current *state* and an associated *reward* as output.
<!-- ML: what are classical approaches below here? -->
In contrast to classical approaches, there are few restrictions on what comprises a state or action. 
The state may be continuous or discrete, completely or partially observed, single or multidimensional.
For instance, we might represent an ecological community of 10 interacting species using an environment with a 10-dimensional continuously valued state space representing the possible population densities of each species.
The action space can be similarly complex.
For the previous environment, we might set the action space to be continuously valued and 10-dimensional, where each dimension corresponds to the amount of conservation effort aimed at each of these species during a time step.   
The main focus of an RL environment, however, is in the environment's transition dynamics and reward function.
The designer of the environment can make the environment follow any transition and reward function provided that both are functions of the current state and inputted action.
This ability allows for RL environments to model a broad range of decision making problems.
For example, we can set the transitions to be deterministic or stochastic.
We can also specify the reward function to be *sparse*, whereby a positive reward is only received for state-action pairs that require a long sequence of actions to come across, e.g. the end point in a maze.
In other environments, an agent may have to learn to forgo immediate reward (or even accept an initial negative reward or cost) in order to maximize the net discounted reward over the simulation period, as we illustrate in examples here.
Another important consideration in RL is whether to use an episodic or non-episodic environment.
In episodic environments, there is a finite number of time steps during which the agent interacts with the environment, and for these environments, the designer must specify a stopping condition.
Conversely, with non-episodic environments, the agent can interact with the environment ad infinitum. 
For this paper, we only consider episodic environments.

@@

>>>>

<!-- something to introduce this section/transition from the intro might be helpful: "We start with a brief overview of RL problems and terminology:" -->

Any application of RL can be divided into two components: an *environment* and an *agent*.
An *environment* is typically a computer simulation, though it is possible to use the real world as the RL environment [@robotics].
Over the course of an episode, an *agent*, which is often a computer program, continuously interacts with the environment. 
At each time step, the agent performs an *action*, and as a result of this action, the environment will determine the next *state* and numerical *reward* that the agent will observe.
The goal of the agent is to learn how to maximize its expected cumulative reward.
The agent learns how to achieve this objective during a period called *training*.
In training, the agent *explores* the available states and actions. 
Once the agent comes across a highly rewarding sequence of states and actions, the agent will reinforce this behavior so that it is more likely for the agent to *exploit* the same high reward trajectory in future episodes.
Throughout this process, the agent's behavior is codified into what is called a *policy*, which is a mapping of states to actions.

<!-- ^ this is great! I do think keeping a table of these terms would be helpful (and refering to that table in the text here) -->

To solve the RL problem, algorithms either take a *model-free* or *model-based* approach. 
The distinction is that *model-free* algorithms do not have access to a model of the environment; yet, *model-based* algorithms use a model of the environment to achieve the RL objective.
A trade-off between these approaches is that when it is possible to quickly learn a model of the environment or the model is already known, model-based algorithms tend to require much less interaction with the environment to learn good-performing policies [@mbpo].
Yet, frequently, learning a model of the environment is very difficult, and in these cases, model-free algorithms tend to outperform [@CITE].
Since most of the recent focus in RL has gone towards model-free algorithms, we will focus on model-free agents for the rest of the paper.

Amongst model-free algorithms, there is divergence around what functions the agent is attempting to learn.
Generally, model-free agents are either learning a policy function, a value function or both.
For context, *value functions* are proxies for how high of a cumulative reward an agent can expect to receive from a given state or state-action pair.
If an agent knows the value function, then the agent can find the optimal action at any state by selecting the action that will maximize the value function in expectation.
The class of algorithms that exclusively try to learn a value function are called *value learning* methods.
By contrast, *Policy gradient* algorithms exclusively learn the policy.
*Actor-critic* algorithms attempt to learn both a policy function and a value function, whereby the value function, which is learned separately from the policy, is used to inform the agent on the goodness of a selected action.
There is a handful of advantages and disadvantages for each of these approaches that we will not go in depth -- for more, see [@CITE]; yet, the main point of distinction comes from the bias-variance trade-off.
Policy gradients tend to struggle with high variance, value-learning tends to be biased while actor-critic methods balance these two concerns.

@@

An environment is a mathematical function, computer program, or real world experience that takes an agent's proposed *action* as input and returns an observation of it's current *state* and associated *reward* as output.
In contrast to classical approaches, such as dynamic programming, there are few restrictions on what comprises a state or action, at least in principle.
The state may be continuous or discrete, completely or only partially observed, one or many dimensions.
For instance, we might represent a ecological community of 10 interacting species using an environment with 10 continuously valued states representing the possible population densities of each species.
The action space may similarly complex, and the response of the environment to any possible action may be deterministic or stochastic in nature.
The reward will in general depend on both the current state and choice of actions.
An environment must also specify a stopping condition, such as when a particular state is reached or a pre-defined interval of time has elapsed.
In some environments, a reward is only received after the agent successfully completes a potentially long sequence of complex actions, such as solving a maze.
In other environments, an agent may have to learn to forgo immediate reward (or even accept an initial negative reward or cost) in order to maximize the net discounted reward over the simulation period, as we illustrate in examples here.

@@

<!-- second pass at this -->

<!-- from end of Carl's discussion edits: Such benchmarks may further improve collaboration across disciplines as well, attracting the attention of AI expertise to these key issues facing our planet. -->
<!-- If RL opens the door to greater role for artificial intelligence and engagement of technology firms in the machinery that determines environmental policy, this raises new concerns as well as new opportunities.   -->

The increasingly influential role of AI in sectors as diverse as policing, finance, and employment has raised concern about the ethics, biases, and power entrenched in both decision-making algorithms and their outputs [@kalluri2020]. 
As algorithms play a greater role in environmental and conservation policy, will similar issues arise?
While this question has garnered  recent interest in the academic literature [@wearn2019; @adams2019; @scoville2021], the use of decision making algorithms in environmental contexts is still relatively isolated and in it's early stages.
Examples of AI applications, and corresponding concerns, have largely been confined to the context of enforcing environmental laws (i.e. algorithms which predict wildlife poaching events [@adams2019] or identify illegal fishing behavior [@mcdonald2021; @swartz2021]). 
However, more diverse applications of AI algorithms - such as applications which contribute to meeting sustainable development goals [@vinuesa2020] or designing  protected area networks - raise additional concerns of entrenching power asymmetries between those reliant on natural resources and those designing algorithms and their objectives.
These issues of power become particularly acute when private industry holds the keys to propriety algorithms or environmental data, and generally operate without the same community engagement obligations of public agencies. 
In these cases, good intentions might not be sufficient to ensure equitable outcomes. 
Interdisciplinary collaborations with experts in algorithmic governance and ethics will be important ensure that applications of AI to ecological challenges promotes equity in both the procedural process of developing objective functions and the distribution of benefits that result from decisions.
